Taking Dyads Seriously

### 2018-02-03 paper/1_intro.tex
some changes and notes

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, International Studies Quarterly published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM) to estimate regression coefficients.  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters, so as to reflect the potential clustering of cases. In practice, these palliatives rarely work because they fail to address the fundamental data generating process that remains a threat to inference. Namely, it is not just the diagonals of the variance-covariance matrix that are affected \citep{beck:2012}.

>>>>

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, \textit{International Studies Quarterly} published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM) to estimate regression coefficients.  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by having non-iid relational data can be addressed by recalculating the standard errors of estimated parameters, so as to reflect the potential clustering of cases. In practice, these palliatives rarely work because they fail to address the fundamental data generating process that remains a threat to inference. Namely, it is not just the diagonals of the variance-covariance matrix that are affected \citep{beck:2012,franzese:hayes:2007,king:roberts:2014}.

### 2018-02-03 paper/2_model.tex
some changes and notes

In working with relational data, scholars typically begin by structuring it as a set of dyadic observations stacked on top of one another. This makes sense if each observation is independent of the others. For example, a conflict sent from the United States to Japan, is assumed to be independent of any action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While the assumption that most begin with is that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause. To uncover the underlying structure of relational data, it is helpful to shift towards restructuring dyadic data in the form of a matrix---often referred to as an adjacency matrix---as shown in Figure~\ref{fig:adjMatDeps}. Rows designate the senders of an event and columns the receivers. The cross-sections in this matrix represent the actions that were sent by an actor in the row to those designated in the columns. Thus $y_{ij}$ designates an action $y$, such as a conflictual event or trade flow, that is sent from actor $i$ to actor $j$.

@@

\begin{figure}[ht]
	\includegraphics[width=.6\textwidth]{stochEquiv_v2.pdf}
	\caption{Visualization of network with meso-scopic features.}
	\label{fig:thirdDeps}
\end{figure}

The visualization illustrates that the actors belonging to the same group have a higher likelihood of having an interaction with each other, whereas interactions across groups are rarer. A prominent example of a network with this type of structure was found by \citet{adamic:glance:2005}, who visualized the ways in which right and left leaning political blogs linked to one another preceding the 2004 United States Presidential Election. \citeauthor{adamic:glance:2005} find that the degree of interaction between right and left leaning blogs was minimal, and that most blogs linked to others that were politically similar. This showcases the types of higher-order dependencies that can emerge in relational data. First, the fact that interactions were determined by a shared attribute, in this case political ideology, is an example of \textit{homophily}. Homophily explains the emergence of patterns such as transitivity (``a friend of a friend is a friend'') and balance (``an enemy of a friend is an enemy''), which also have a long history in international relations. The other major type of meso-scopic feature that emerges in relational data is community structure \citep{mucha:etal:2010}, which is often formalized through the concept of stochastic equivalence \citep{anderson:etal:1992}. This concept simply refers to the idea that groups of nodes that act similarly in the network are stochastically equivalent. In the example we have laid out above each of the left leaning blogs would be considered stochastically equivalent to one another.

>>>>

In working with relational data, scholars typically begin by structuring it as a set of dyadic observations stacked on top of one another. This makes sense if each observation is independent of the others. For example, a conflict sent from the United States to Japan, is assumed to be independent of any action that Japan may send to the United States. Additionally, every action sent by Japan to others in the system is considered independent even though each of those interactions involves a common sender, i.e, Japan. While the assumption that most begin with is that each dyadic interaction is taking place in isolation of the others, we know this assumption to be false in theory and in practice. Relational data comes with an explicit structure that generally leads to particular types of dependencies. The importance of accounting for the underlying structure of our data has been a lesson well understood, at least, when it comes to time-series cross-sectional data (TSCS) within political science \citep{beck:katz:1995,beck:etal:1998}. As a result, it is now standard practice to take explicit steps to account for the complex data structures that emerge in TSCS applications and the unobserved heterogeneity that they cause.

To uncover the underlying structure of relational data, it is helpful to shift towards restructuring dyadic data in the form of a matrix---often referred to as an adjacency matrix---as shown in Figure~\ref{fig:adjMatDeps}. Rows designate the senders of an event and columns the receivers. The cross-sections in this matrix represent the actions that were sent by an actor in the row to those designated in the columns. Thus $y_{ij}$ designates an action $y$, such as a conflictual event or trade flow, that is sent from actor $i$ to actor $j$. 
% add a note about undirected data sharing a similar structure as well

@@

\begin{figure}[ht]
	\includegraphics[width=.6\textwidth]{stochEquiv_v2.pdf}
	\caption{Visualization of a network with meso-scopic features.}
	% need a simpler caption here
	\label{fig:thirdDeps}
\end{figure}

The visualization illustrates that the actors belonging to the same group have a higher likelihood of having an interaction with each other, whereas interactions across groups are rarer. A prominent example of a network with this type of structure is discussed by \citet{adamic:glance:2005}, who visualizes the ways in which right and left leaning political blogs linked to one another preceding the 2004 United States Presidential Election. \citeauthor{adamic:glance:2005} find that the degree of interaction between right and left leaning blogs was minimal, and that most blogs linked to others that were politically similar. This showcases the types of higher-order dependencies that can emerge in relational data. First, the fact that interactions were determined by a shared attribute, in this case political ideology, is an example of \textit{homophily}. Homophily explains the emergence of patterns such as transitivity (``a friend of a friend is a friend'') and balance (``an enemy of a friend is an enemy''), which also have a long history in international relations. The other major type of meso-scopic feature that emerges in relational data is community structure \citep{mucha:etal:2010}, which is often formalized through the concept of stochastic equivalence \citep{anderson:etal:1992}. This concept simply refers to the idea that groups of nodes that act similarly in the network are stochastically equivalent. In the example we have laid out above each of the left leaning blogs would be considered stochastically equivalent to one another.

### 2018-02-03 paper/netsMatter.tex
fix source of compilation error
some changes and notes

\small{\singlespacing{
\begin{abstract}
	%International relations scholarship is filled with dyads. This includes dyadic hypotheses and propositions, but especially dyadic data. Yet, much of the work conducted brushes aside the statistical complications that come with analyzing relational data, namely that the observations do not satisfy the conditional independence criterion. As a result, these studies often produce results with biased coefficient estimates and poorly calibrated standard errors, factors which have consequences for hypothesis testing. We present an alternative regression based approach that accounts for the dependencies that complicate inferential analysis in the context of relational data using a set of additive and multiplicative effects within a Bayesian framework. To highlight the utility of this framework for studying relational data, we present a simulation exercise highlighting the model's ability to account for the dependencies that emerge in relational data. Additionally, we replicate five prominent studies in recent IR scholarship and compare the standard approach to the latent factor approach. For each study, our approach shows that conventional methods overstate the effect of key variables in the study of interstate conflict, underestimate the uncertainty in these effects, and in some cases lead researchers to faulty conclusions about the statistical significance of their variables. Further, the new approach dominates in terms of out-of-sample cross-validations, rendering it more useful in forecasting applications and in modeling the data generating process behind outcomes of interest.

	Much of international relations scholarship concerns dyads: dyadic hypotheses and especially dyadic data. Yet, contemporary research brushes aside the complications that come with analyzing interdependent relational data. Dyadic observations do not typically satisfy the conditional independence criterion required for many statistical approaches. As a result, many studies often produce results with biased coefficient estimates and poorly calibrated standard errors. These biases have profound consequences for evaluating parametric models. We present an alternative, regression-based, approach that accounts for the dependencies complicating this type of analysis. We first present a simulation exercise highlighting the model's ability to account for the dependencies that emerge in relational data. In addition, we replicate five prominent studies in recent international relations scholarship, comparing the standard approach to our alternative. For each study, we find that conventional methods overstate the effect of key variables in the study of interstate conflict (and trade), underestimate the uncertainty in these effects, and in some cases lead researchers to faulty conclusions about the statistical significance and substantive importance of their variables. Further, we show that our approach dominates in terms of out-of-sample cross-validations, rendering it more useful in forecasting applications and in modeling the data generating process behind outcomes of interest.
\end{abstract}
}}

>>>>

\small{\singlespacing{
\begin{abstract}
	Much of international relations scholarship concerns dyads: dyadic hypotheses and especially dyadic data. Yet, contemporary research brushes aside the complications that come with analyzing interdependent relational data. Dyadic observations do not typically satisfy the conditional independence criterion required for many statistical approaches. As a result, many studies often produce results with biased coefficient estimates and poorly calibrated standard errors. These biases have profound consequences for evaluating parametric models. We present an alternative, regression-based, approach that accounts for the dependencies complicating this type of analysis. We first present a simulation exercise highlighting the model's ability to account for the dependencies that emerge in relational data. In addition, we replicate five recent studies in recent international relations scholarship, comparing the standard approach to our alternative. For each study, we find that conventional methods overstate the effect of key variables in the study of interstate conflict (and trade), underestimate the uncertainty in these effects, and in some cases lead researchers to faulty conclusions about the statistical significance and substantive importance of their variables. Further, we show that our approach dominates in terms of out-of-sample cross-validations, rendering it more useful in forecasting applications and in modeling the data generating process behind outcomes of interest.
\end{abstract}
}}
